---
title: "Getting started: a start-to-finish dose-response walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: a start-to-finish dose-response walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This vignette walks the Ward rotavirus human-challenge data from raw counts to a reported risk number, interpreting every output along the way. It answers two questions an end user actually has: *what does my data need to look like going in*, and *what are the results saying coming out*.

```{r setup}
library(qrmavfar)
```

## Load

The analysis needs **three columns, one row per dose group**: the administered `dose`, and the numbers of subjects who did (`positive`) and did not (`negative`) respond. The package ships this exact shape as `ward_rotavirus`:

```{r load-data}
data(ward_rotavirus)
ward_rotavirus
```

There are two ways in. The **in-memory** path standardizes a data frame you already have:

```{r as-dose-response}
ward <- as_dose_response(ward_rotavirus)
ward
```

The **from-disk** path reads the same data from the bundled text file — this is what you would use for your own delimited file:

```{r read-from-disk}
ward_path <- system.file("extdata", "Ward_rotavirus.txt", package = "qrmavfar")
ward_from_file <- read_dose_response(ward_path)

identical(ward, ward_from_file)
```

Both routes produce the standardized `dose · positive · negative · total · response` tibble, where `total = positive + negative` and `response = positive / total`. Column names are auto-detected case- and punctuation-insensitively (`pos`, `positive_response`, etc.); override detection with `as_dose_response(data, dose =, positive =, negative =)`. Rows sharing a dose are summed. Fitting requires **at least 3 distinct doses** and **more than 1 total positive response** — non-conforming data is rejected with an explanatory error.

## Run

`analyze_dose_response()` does everything in one call: trend test, model fits, diagnostics, comparison, assessment, and bootstrap uncertainty. A fixed `seed` makes the bootstrap reproducible.

```{r analyze}
analysis <- analyze_dose_response(
  ward,
  bootstrap_times = 1000, # reduced for a fast vignette build; the default is 10000
  resample = "observed",
  seed = 2026
)

analysis
```

The rest of this vignette unpacks each component of that `qdr_analysis` object.

## Interpret each output

### Trend — is there a dose-response signal at all?

```{r trend}
analysis$trend
```

`dose_trend_test()` runs a one-sided log-dose trend test. `passes = TRUE` means a **monotonic increasing** dose-response relationship was detected — response rises with dose.

> **What this is really saying:** if `passes` were `FALSE`, there is no upward dose-response signal to model, and fitting mechanistic curves would not be warranted. Here it passes, so we proceed.

### Goodness of fit — does each model fit the data at all?

```{r goodness}
analysis$goodness_of_fit
```

This is the **absolute** check. Each model's `deviance` is compared against a chi-squared `critical_value`; `good_fit` is `TRUE` when the deviance falls below it (equivalently, `p_value` \> 0.05). `assessment` records `adequate` or `inadequate`.

> **What this is really saying:** does this model describe the data acceptably on its own terms? A model can "win" a comparison and still fit badly — this is the gate that catches that.

### Comparison — which model wins, and did the extra parameter earn its keep?

```{r comparison}
analysis$comparison
```

This is the **relative** check. The two-parameter Beta-Poisson is tested against the one-parameter Exponential (its nested limiting case) via the `deviance_difference` on the extra degree of freedom, alongside `AIC`/`BIC` and their deltas. `significant_improvement` and `preferred` flag the winner.

> **What this is really saying:** the Beta-Poisson always fits at least as well because it has an extra parameter — `significant_improvement` asks whether that extra flexibility bought a *statistically real* improvement, or just curve-fit noise.

### Assessment — the headline verdict

```{r assessment}
analysis$assessment[, c("model", "appropriate", "preferred", "recommendation", "conclusion")]
```

`build_model_assessment()` fuses the absolute and relative checks into one `recommendation` label per model:

-   `recommended` — fits adequately **and** is preferred.
-   `acceptable_alternative` — fits adequately but is not preferred.
-   `preferred_but_inadequate` — preferred but does **not** fit adequately (a red flag: the "best" model is still wrong).
-   `not_recommended` — neither.

> **What this is really saying:** this is the one row a risk assessor reads first. `recommended` means "use this model with confidence"; the `conclusion` column spells it out in plain language.

### Consensus — an experimental cross-check

```{r consensus}
consensus_model_decision(analysis)
```

`consensus_model_decision()` treats the chi-squared, AIC, and BIC selections as three votes and reports the `agreement` level (`unanimous`, `majority`, or `no_consensus`).

> **What this is really saying:** an experimental sanity check on the relative selection. It does **not** replace the absolute goodness-of-fit gate above — a unanimously-voted model can still fit inadequately.

### Bootstrap intervals — the final risk numbers and their uncertainty

```{r bootstrap-ci}
bootstrap_confint(analysis$bootstraps$beta_poisson, levels = c(0.95))
```

The bootstrap refits the model to resampled data many times and reports percentile confidence intervals — not just for the parameters (`alpha`, `n50`) but for the **effective doses** `ed10` and `ed50`: the doses producing a 10% and 50% response.

> **What this is really saying:** these are the deliverable. `ed50` is the median infectious dose; its interval is the honest uncertainty around it, driven by how much data you have and how noisy it is.

## Plots

`autoplot()` on the analysis shows each fitted curve over the observed points:

```{r plot-analysis}
ggplot2::autoplot(analysis)
```

`autoplot()` on a bootstrap object shows the sampling distribution of the fitted parameters — the visual form of the intervals above:

```{r plot-bootstrap}
ggplot2::autoplot(analysis$bootstraps$beta_poisson)
```

## The final answer

Read the recommended model's `ed50` and its interval, and you have the number a risk assessor reports:

```{r final-answer}
# Which model did the assessment recommend?
recommended <- analysis$assessment$model[analysis$assessment$recommendation == "recommended"][1]
recommended

# Its bootstrap intervals, then keep only the ed50 row.
ci <- bootstrap_confint(analysis$bootstraps[[recommended]], levels = c(0.95))
ci[ci$term == "ed50", ]
```

That single row — the median infectious dose of the recommended model with its 95% percentile interval — is the deliverable: *the dose at which half of exposed subjects are expected to be infected, with a defensible bound on how sure we are.*